Think deep for reef resilience

### 2025-07-23 paper/paper_draft.qmd
Further information about connectivity methods and possible improvements

@@

It is a common conclusion in coral reef ecology that a primary facet of reef resilience is
the degree of larval connectivity between reefs and the supply of recruits that a reef
receives following disturbance [@hock2017] **expand citations**. In contrast, our results
suggest that there are few differences between resilient and non-resilient reefs in
connectivity metrics such as the ratio of outgoing to incoming larval connectivity and total
connectivity strength. Furthermore, our results suggest no relationship between connectivity
and the ability of reefs to maintain positive carbonate budgets under future conditions.
There has been contrasting evidence [@bode2018; @hock2017] surrounding whether recent
generation connectivity modelling is able to identify reefs that are important for recovery
processes. A key separation of these investigations is their use of differing larval
connectivity models. Reef depth was not considered in these investigations of reef network
resilience. These investigations were performed using prior generations of larval
connectivity models parameterised for 2008-2013 [@hock2017] and 1996-2002 [@bode2018] time
periods. Using updated 1km resolution larval connectivity across the whole GBR, with a
limited temporal sample of particle flows, our results suggest that larval connectivity
alone is not as important for reef resilience and recovery as environmental refugia from
thermal stress such as depth. Our findings may differ due to the highly uncertain nature of
connectivity methods used **citation**, the way larval flows and juvenile recruitment are
implemented in ADRIA - CoralBlox, and the magnitude of early bleaching events decreasing the
breeding capacity of strongly connected reefs. Further improvements and research in
connectivity modelling are necessary to provide a robust comparison of all factors
influencing reef resilience.

>>>>

Twenty-four total days of connectivity data were used in this investigation, 2015, 2016,
2018 and 2019 are represented represented by 3 daily connectivity matrices each, while 2017
and 2020 are represented with 5 daily matrices each. Connectivity levels are implemented in
ADRIA - CoralBlox as mean over the 24 days for each location.

@@

It is a common conclusion in coral reef ecology that a primary facet of reef resilience is
the degree of larval connectivity between reefs and the supply of recruits that a reef
receives following disturbance [@hock2017; @mumby2008; @torres2024]. In contrast, our
results suggest that there are few differences between resilient and non-resilient reefs in
connectivity metrics such as the ratio of outgoing to incoming larval connectivity and total
connectivity strength. Furthermore, our results suggest no relationship between connectivity
and the ability of reefs to maintain positive carbonate budgets under future conditions.
There has been contrasting evidence [@bode2018; @hock2017] surrounding whether recent
generation connectivity modelling is able to identify reefs that are important for recovery
processes, with findings dependent on the larval connectivity model used. These
investigations were performed using prior generations of larval connectivity models
parameterised for 2008-2013 [@hock2017] and 1996-2002 [@bode2018] time periods. Using
updated 1km resolution larval connectivity across the whole GBR, with a limited temporal
sample of particle flows, our results suggest that larval connectivity alone is not as
important for reef resilience and recovery as environmental refugia from thermal stress such
as depth.

Our findings on the importance of connectivity may differ due to the highly uncertain nature
of larval connectivity modelling [@vogt-vincent2023], the way larval flows and juvenile
recruitment are implemented in ADRIA - CoralBlox, and the magnitude of early bleaching
events decreasing the breeding capacity of strongly connected reefs. Analysing daily
connectivity data Vogt-Vincent et al. [-@vogt-vincent2023] suggest that time-mean
connectivity may not be suitable for capturing the seasonal - ecological time scale,
connectivity between locations due to the high daily variability in connectivity data. This
is supported by the connectivity data used in our investigation, which have large relative
standard deviations for connectivity between reefs over the 24 days used **figure**. Using
alternative summarisation methods that capture ecological-timescale patterns in
coral larval dispersal, such as cumulated implicit connectivity [@vogt-vincent2023],
future investigations may find increased importance of both depth and reef connectivity for
resilience, in line with [@hock2017; @torres2024]. Further improvements and research in
connectivity modelling are necessary to provide a robust comparison of all factors
influencing reef resilience.
